Finiteness in Cubical Type Theory

### 2020-06-28 sections/finiteness-predicates.tex
fixed equational reasoning syntax
bot more organisation of the split enumerable section

\chapter{Finiteness Predicates} \label{finiteness-predicates}
In this section, we will define and briefly describe each of the five predicates
in Figure~\ref{finite-classification}.
We will also explain \emph{why} there are five separate predicates: how can it
be the case that so many different things describe ``finiteness''?
As we will see, some predicates are too informative (they
tell us more about the underlying type other than it just being finite), or too
restrictive (they don't allow certain finite types to be classified as finite).
These diversions won't be dead-ends, however: the final predicate we will land
on as the ``correct'' (or, more accurately, most useful) notion of finiteness
will be built out of all of the others.
\section{Split Enumerability} \label{split-enumerability}
We will start with a simple notion of finiteness, called \emph{split}
enumerability.

\begin{definition}[Split Enumerable Set] \label{split-enum-def}
  To say that some type \(A\) is split enumerable is to say that there is a list
  \(\mathit{support} : \AgdaDatatype{List}(A)\) such that any value \(x : A\) is in
  \(\mathit{support}\).
  \begin{agdalisting} \label{split-enum-def-eqn}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable/Container.tex]{split-enum-def}
  \end{agdalisting}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  An equivalent version of this predicate was called \verb+Listable+ in
  \cite{firsovDependentlyTypedProgramming2015}.
\end{definition}

\subsection{Terms used in the Definition of Split Enumerability}
We have used two types there which we have not yet defined:
\(\AgdaDatatype{List}\) and \(\AgdaDatatype{\ensuremath{\in}}\).
We will define them here.

@@

Membership also makes more sense when described concretely in terms of lists.
Understood this way, \(x\;\AgdaDatatype{\ensuremath{\in}}\;\mathit{xs}\) means
``there is an index into \(\mathit{xs}\) such that the index points at an item
equal to \(x\)''.
\todo{Probably should put back in some instances here, i.e. bool and top.
  Obviously not the full closure stuff, but one or two instances would be
  instructive.}
\subsection{Derivations}
We have a way to construct finiteness proofs, and a semiring-like toolbox to
combine them.
What we're now interested in is what we can \emph{derive} from them.

First, we will look at how this predicate relates to more traditional, classical
notions of finiteness.
in a classical setting we likely wouldn't mention ``lists'' or the like, and
would instead define finiteness based on the existence of some injection or
surjection.
As it turns out, our definition of finiteness here is precisely the same as the
surjection-based one, in quite a deep way!

First, we will need to define our terms: in HoTT, surjections are a little more
complex than what you'd find in either MLTT or classical mathematics.
\begin{definition}[Split Surjections] \label{split-surjections}
  We define \emph{split} surjections here \cite[definition
  4.6.1]{hottbook}. \todo{To Agda}
  \begin{alignat}{3}
    &\text{sp-surj}(f)          &&\coloneqq \Pi(y : B) , \text{fib}_f(y) \label{sp-surj-eqn} \\
    &A \twoheadrightarrow! \; B &&\coloneqq \Sigma (f : A \rightarrow B) , \text{sp-surj}(f) \label{sp-surj-arrow-eqn}
  \end{alignat}
\end{definition}
Over sets, the surjections and split surjections are the same thing, but there
is a difference one we involve non-set types like the circle.

We will now see that split enumerability is in fact a split surjection in
another form:
\begin{lemma} \label{split-enum-is-split-surj}%
  A proof of split enumerability is equivalent to a split surjection from a
  finite prefix of the natural numbers.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj-type}
  \end{agdalisting}
\end{lemma}
\todo{Use standard equational reasoning syntax}
\begin{proof} \let\qed\relax
  \begin{minipage}[t]{.7\textwidth}\vspace{-\baselineskip}
    \begin{agdalisting*}
      \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
    \end{agdalisting*}
  \end{minipage}
  \begin{minipage}[t]{.19\textwidth} \setstretch{1.1}

      Def.~\ref{split-enum-def} (\AgdaDatatype{\ensuremath{\mathcal{E}!}}) \\
      Eqn.~\ref{container-membership} (\AgdaDatatype{\ensuremath{\in}}) \\
      Eqn.~\ref{sp-surj-eqn}  \\
      Def.~\ref{List} (\(\AgdaDatatype{List}\)) \\
      Eqn.~\ref{container-interp}  \\
      Reassociation \\
      Eqn.~\ref{sp-surj-arrow-eqn}
  \end{minipage}
\end{proof}
In the above proof syntax the
\(\AgdaDatatype{\ensuremath{\equiv \langle{} \rangle{} }}\) connects lines which
are definitionally equal, i.e. they are ``obviously'' equal from the type
checker's perspective.
Clearly, only one line isn't a definitional equality: 
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)

Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{definition}[Injections]
  Injective functions are more straightforward to define constructively than
  surjective ones:
  \begin{alignat}{3}
    & \text{injective}(f) &&\coloneqq \Pi (x, y : A) , f \; x \equiv f \; y \rightarrow x \equiv y \\
    & A \rightarrowtail B &&\coloneqq \Sigma(f : A \rightarrow B) , \text{injective}(f)
  \end{alignat}
\end{definition}
\begin{lemma} \label{split-surj-to-inj}
  A split-surjection from \(A\) to \(B\) implies an injection from \(B\) to
  \(A\).
  \begin{equation}
    (A \twoheadrightarrow! \; B) \rightarrow (B \rightarrowtail A)
  \end{equation}
\end{lemma}
\begin{lemma} \label{inj-discrete}
  For any type \(A\) which injects into a discrete type \(B\), \(A\) is
  discrete.
  \begin{equation}
    \frac{
      A \rightarrowtail B \; \; \; \text{Discrete}(B)
    }{
      \text{Discrete}(A)
    }
  \end{equation}
\end{lemma}

>>>>

\chapter{Finiteness Predicates} \label{finiteness-predicates}
In this section, we will define and briefly describe each of the five predicates
in Figure~\ref{finite-classification}.
We will also explain \emph{why} there are five separate predicates: how can it
be the case that so many different things describe ``finiteness''?
As we will see, some predicates are too informative (they
tell us more about the underlying type other than it just being finite), or too
restrictive (they don't allow certain finite types to be classified as finite).
These diversions won't be dead-ends, however: the final predicate we will land
on as the ``correct'' (or, more accurately, most useful) notion of finiteness
will be built out of all of the others.
\section{Split Enumerability} \label{split-enumerability}
We will start with a simple notion of finiteness, called \emph{split}
enumerability.
This predicate is perhaps the first definition of ``finite'' that someone might
come up with (it's certainly the most common in dependently-typed programming):
put simply, a split enumerable type is a type for which all of its elements can
be listed.
\begin{definition}[Split Enumerable Set] \label{split-enum-def}
  To say that some type \(A\) is split enumerable is to say that there is a list
  \(\mathit{support} : \AgdaDatatype{List}(A)\) such that any value \(x : A\) is in
  \(\mathit{support}\).
  \begin{agdalisting} \label{split-enum-def-eqn}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable/Container.tex]{split-enum-def}
  \end{agdalisting}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  An equivalent version of this predicate was called \verb+Listable+ in
  \cite{firsovDependentlyTypedProgramming2015}.
\end{definition}

Simple as it is, this predicate is arguably the most useful of the finiteness
predicates.
We will see presently why it doesn't entirely capture what we mean by the term
``finite'', but we will nonetheless use it throughout the paper.
Before we dive in to exploring the predicate itself, though, we will need to
explain some of the terms we used in its definition.
\subsection{Terms used in the Definition of Split Enumerability}
We have used two types there which we have not yet defined:
\(\AgdaDatatype{List}\) and \(\AgdaDatatype{\ensuremath{\in}}\).
We will define them here.

@@

Membership also makes more sense when described concretely in terms of lists.
Understood this way, \(x\;\AgdaDatatype{\ensuremath{\in}}\;\mathit{xs}\) means
``there is an index into \(\mathit{xs}\) such that the index points at an item
equal to \(x\)''.
\subsection{Split Surjections}
We will now look at how this predicate relates to more traditional, classical
notions of finiteness.
In a classical setting we likely wouldn't mention ``lists'' or the like, and
would instead define finiteness based on the existence of some injection or
surjection.
As it turns out, our definition of finiteness here is precisely the same as the
surjection-based one, in quite a deep way!

First, we will need to define our terms: in HoTT, surjections are a little more
complex than what you'd find in either MLTT or classical mathematics.
\begin{definition}[Split Surjections] \label{split-surjections}
  We define \emph{split} surjections here \cite[definition
  4.6.1]{hottbook}.
  \begin{agdalisting} \label{sp-surj-eqn}
    \ExecuteMetaData[agda/Function/Surjective/Base.tex]{split-surjective}
  \end{agdalisting} \vspace{-1.5\baselineskip}
  \begin{agdalisting} \label{sp-surj-arrow-eqn}
    \ExecuteMetaData[agda/Function/Surjective/Base.tex]{split-surjection}
  \end{agdalisting}
\end{definition}
Over sets, the surjections and split surjections are the same thing, but there
is a difference one we involve non-set types like the circle.

We will now see that split enumerability is in fact a split surjection in
another form:
\begin{lemma} \label{split-enum-is-split-surj}%
  A proof of split enumerability is equivalent to a split surjection from a
  finite prefix of the natural numbers.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj-type}
  \end{agdalisting}
\end{lemma}
\begin{proof} \let\qed\relax \vspace{-1\baselineskip}
  \begin{minipage}[t]{.7\textwidth}\vspace{-\baselineskip}
    \begin{agdalisting*}
      \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
    \end{agdalisting*}
  \end{minipage}
  \begin{minipage}[t]{.19\textwidth} \setstretch{1.1}

      Def.~\ref{split-enum-def} (\AgdaDatatype{\ensuremath{\mathcal{E}!}}) \\
      Eqn.~\ref{container-membership} (\AgdaDatatype{\ensuremath{\in}}) \\
      Eqn.~\ref{sp-surj-eqn}  \\
      Def.~\ref{List} (\(\AgdaDatatype{List}\)) \\
      Eqn.~\ref{container-interp}  \\
      Reassociation \\
      Eqn.~\ref{sp-surj-arrow-eqn}
  \end{minipage}
\end{proof}
In the above proof syntax the
\(\AgdaDatatype{\ensuremath{\equiv \langle{} \rangle{} }}\) connects lines which
are definitionally equal, i.e. they are ``obviously'' equal from the type
checker's perspective.
Clearly, only one line isn't a definitional equality: 
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)
\subsection{Instances}
We have now characterised the informativeness of split enumerability (it's the
same as a split surjection), so we should also look at its
\emph{restrictiveness}.
For this we'll first look at some simple types which are clearly finite, and
we'll see how to prove that's the fact using split enumerability.
\begin{lemma}
  \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are split enumerable.
\end{lemma}
\begin{proof}
  These three types are quite obviously finite: we will show only the proof of
  finiteness for \(\AgdaDatatype{Bool}\) here for brevity's sake.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \end{agdalisting}
\end{proof}
\subsection{Decidable Equality}
Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{definition}[Injections]
  Injective functions are more straightforward to define constructively than
  surjective ones:
  \begin{alignat}{3}
    & \text{injective}(f) &&\coloneqq \Pi (x, y : A) , f \; x \equiv f \; y \rightarrow x \equiv y \\
    & A \rightarrowtail B &&\coloneqq \Sigma(f : A \rightarrow B) , \text{injective}(f)
  \end{alignat}
\end{definition}
\begin{lemma} \label{split-surj-to-inj}
  A split-surjection from \(A\) to \(B\) implies an injection from \(B\) to
  \(A\).
  \begin{equation}
    (A \twoheadrightarrow! \; B) \rightarrow (B \rightarrowtail A)
  \end{equation}
\end{lemma}
\begin{lemma} \label{inj-discrete}
  For any type \(A\) which injects into a discrete type \(B\), \(A\) is
  discrete.
  \begin{equation}
    \frac{
      A \rightarrowtail B \; \; \; \text{Discrete}(B)
    }{
      \text{Discrete}(A)
    }
  \end{equation}
\end{lemma}

### 2020-06-28 sections/topos.tex
bot more organisation of the split enumerable section

\section{Categories in HoTT}
\section{Closure}
\todo{This is just the closure bits pulled over from the finiteness predicates
  section. They need to be reworked to fit in here.}
\subsection{Split Enumerability Closure}
Now that we have a suitable definition of finiteness, we will next prove that
some things are finite.
\begin{lemma}
  \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are split enumerable.
\end{lemma}
\begin{proof}
  These three types are quite obviously finite: we will show only the proof of
  finiteness for \(\AgdaDatatype{Bool}\) here for brevity's sake.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \end{agdalisting}
\end{proof}
With the most basic simple types out of the way, the obvious next choice is the
(non-dependent) sums and products: \(\uplus\) and \(\times\).
Both of these types can be constructed from the \emph{dependent} sum, however,
so that is the type we will prove finite.
From that we can derive a much wider array of finiteness proofs.

>>>>

\section{Categories in HoTT}
\section{Closure}
\todo{This is just the closure bits pulled over from the finiteness predicates
  section. They need to be reworked to fit in here.}
\subsection{Split Enumerability Closure}
Now that we have a suitable definition of finiteness, we will next prove that
some things are finite.
With the most basic simple types out of the way, the obvious next choice is the
(non-dependent) sums and products: \(\uplus\) and \(\times\).
Both of these types can be constructed from the \emph{dependent} sum, however,
so that is the type we will prove finite.
From that we can derive a much wider array of finiteness proofs.
